---
title: "Data story - 100 meters"
author: "Adrian Meier"
date: "February 7, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Choice of technology
During the first interview David mentioned that the internship project is currently based on R. Having no prior experience with R I decided to take the learning opportunity and got to work with R and RMarkdown employing the libraries: rvest, dplyr, stringr, ggplot2, lubridate and ExtDist.

## Choosing a starting point
The 100 meters Wikipedia page <https://en.wikipedia.org/wiki/100_metres> is full of data tables. A total of 17 tables rank racers in all kinds of categories. The '100 meters per age category' section caught my eye. It lists the best performances of athletes between 5 and 19 years of age. What sparked my interest is the 'Age' column. The record holders aged five are both less than two weeks away from their sixth birthday. Clearly the age has a significant impact on performance at an early age. As a first question, I want to investigate the correlation of the athletes age (within an age category) to their absolute age. 

## Scraping wikipedia
Lets begin by reading the two tables of the *100 meters per age category* section. I wrote a function using the rvest library to read a HTML table given the corresponding XPATH from the wiki URL once. The table is stored locally as a csv file and loaded locally if available. Then I call the function with the arguments for the tables of interest.
```{r global, echo=FALSE, message=FALSE}
# install.packages("rvest")
# install.packages("dplyr")
# install.packages("stringr")
library(rvest)
library(dplyr)
library(stringr)
url = "https://en.wikipedia.org/wiki/100_metres#100_metres_per_age_category"
```

```{r load_data, echo=TRUE, message = FALSE}
# load data from file directly if already scraped data and scrape and write csv file otherwise
load_data = function(data_name, data_xpath){
  data_filepath = paste("~/sprint/data/", data_name, ".csv", sep = "")
  # check if data file exists
  if (!file.exists(data_filepath)){
    # access data at url by XPATH
    loaded_data = url %>% 
      read_html() %>% 
      html_node(xpath = data_xpath) %>% 
      html_table(fill = TRUE)
    # export to csv
    write.csv(loaded_data, file=data_filepath, row.names = FALSE)
  } else {
    loaded_data = read.csv(data_filepath, check.names = FALSE)
  }
  return(loaded_data)
}
```

```{r load_data call, echo=FALSE, message=FALSE}
PAB_name = "per_age_boys"
PAB_xpath = "/html/body/div[3]/div[3]/div[5]/div[1]/div[9]/div[11]/table/tbody/tr/td[1]/table"

PAG_name = "per_age_girls"
PAG_xpath = "/html/body/div[3]/div[3]/div[5]/div[1]/div[9]/div[11]/table/tbody/tr/td[2]/table"

per_age_boys = load_data(PAB_name, PAB_xpath)
per_age_girls = load_data(PAG_name, PAG_xpath)
glimpse(per_age_girls)
```

Next I do some post processing on the tables. To prevent ambiguity I rename the first 'Age' column to 'Age_category'. Further I drop columns that I won't need, add a gender flag and then combine the data in a single table.

```{r per_age postprocessing, echo=FALSE, message=FALSE}
# rename ambiguous column
colnames(per_age_boys)[1] = "Age_category"
colnames(per_age_girls)[1] = "Age_category"

# construct one table with age categories, containing boys and girls ages at time of the corresponding records
PAB_table = select(per_age_boys, Age_category, Athlete, Age)
PAB_table$Gender = "male"
PAG_table = select(per_age_girls, Age_category, Athlete, Age)
PAG_table$Gender = "female"

# combine into one table
age_category_table = rbind(PAB_table, PAG_table)
glimpse(age_category_table)
```
Now I add a column which will store the days until the athletes birthday. This shows how many (or few) days were remaining for the athlete to compete in this age category (and potentially get a chance to improve their time). To calculate the remaining days, I have to compare the age (given as "X years, Y days") with the age category. Maybe there are neat functions which can handle this data format... I solved it with some stringr regexp magic.
```{r calc_days_remaining, echo=FALSE, message=FALSE}
age_category_table$Days_remaining = NA
# calculate remaining days where athlete is eligible to run in the age category where they set the corresponding record
for (idx in 1:length(age_category_table$Age_category)){
  age_days = as.numeric(str_extract(strsplit(toString(age_category_table$Age[idx]), ",")[[1]][2], "\\d+"))
  age_category_table$Days_remaining[idx] = 365 - age_days
  age_years = as.numeric(str_extract(strsplit(toString(age_category_table$Age[idx]), ",")[[1]][1], "\\d+"))
  # if more than a year younger than age category
  if (age_years < age_category_table$Age_category[idx]){
    years_diff = age_category_table$Age_category[idx] - age_years
    age_category_table$Days_remaining[idx] = age_category_table$Days_remaining[idx] + 365*(years_diff)
  }
}
```
Having done that, it's time for the first plot!
```{r first age category plot, echo=FALSE, message=TRUE}
# graphic representing the time and age, with anomaly explanaition.
age_category_table$Gender = factor(age_category_table$Gender)
plot(age_category_table$Age_category, age_category_table$Days_remaining, 
     col=age_category_table$Gender,
     pch = 17,
     xlab = "Age category",
     ylab = "Days remaining",
     main = "Age category records set with #days remaining within category")
legend("topleft", legend=levels(age_category_table$Gender), col=c(1:2), pch = 17)
```  

Surprisingly there's one outlier: Payton Payne's record set at age 11 improved all previous times run by girls age 12. A closer look shows that Payton holds records in five categories.
```{r outlier, echo=FALSE, message=FALSE}
outlier_per_age_girls = select(filter(per_age_girls, Athlete == "Payton Payne"), Age_category, Athlete, Date, Age)
print(outlier_per_age_girls)
```
Payton set her last record 28 July 2019. I assume in 2020 there were close to no competitions due to covid. I think it's fair to assume she would have further improved the age category 12 record had she gotten the chance. Therefore I decided to remove the outlier and to not count her outstanding performance in the age 12 categories.

Additionally I plot a half year horizontal line in purple and a least squares linear regression line in blue:
```{r removing outlier and final age category plot, echo=FALSE, message=FALSE}
# remove outlier and analyze (fit, avg,...) add half year avg. line
outlier_row = which(age_category_table$Days_remaining > 365)
age_category_table = age_category_table[-c(outlier_row),]

# graphic representing the time and age, with anomaly explanaition.
age_category_table$Gender = factor(age_category_table$Gender)
plot(age_category_table$Age_category, age_category_table$Days_remaining, 
     col=age_category_table$Gender,
     pch = 17,
     xlab = "Age category",
     ylab = "Days remaining",
     main = "Enhanced age category plot", 
     ylim = c(0, 365))
legend("topleft", legend=levels(age_category_table$Gender), col=c(1:2), pch = 17)

# half a year line
abline(h = 365/2, col = "purple")
# linear regression
lr_age_category = lm(Days_remaining ~ Age_category, data = age_category_table) 
abline(a = coef(lr_age_category)[1], b = coef(lr_age_category)[2], col="blue")
```  

The linear regression (blue) shows nicely what one expects: While age has a big impact on the performance of young athletes, the impact is lost over time. It seems that young athletes improve their performance in close relation to getting older, while from maybe 16 onward the age factor appears to lose it's weight.

## Age and adult athletes best performance
This prompted the follow up question: At what age did the all-time fastest athletes run their record times?

To answer this question I load the tables from the Wikipedia page sections 'All-time top 25' for men and women. I make sure to remove duplicate entries of record holders who ran their record time more than once.

```{r load and postprocess ATB, echo=FALSE, message=FALSE}
# xpath to all time top 25 athletes.
ATBW_name = "all_time_best_women"
ATBW_xpath = "/html/body/div[3]/div[3]/div[5]/div[1]/div[9]/table[2]"

ATBM_name = "all_time_best_men"
ATBM_xpath = "/html/body/div[3]/div[3]/div[5]/div[1]/div[9]/table[1]"

all_time_best_women = load_data(ATBW_name, ATBW_xpath)
all_time_best_men = load_data(ATBM_name, ATBM_xpath)

# clean tables - keeping every athlete exactly once.
all_time_best_women = select(all_time_best_women[-c(26,27),], Athlete, Date)
all_time_best_men = select(filter(all_time_best_men, !grepl("\\d", all_time_best_men$Athlete)), Athlete, Date)
```
To my disappointment, while the date when the record was run is present in the table, the age of the athlete's is missing. I decide to collect the links to the athlete's personal Wikipedia page, to then extract the age information from there. I use the following functions to first collect all links. This includes links to the cities where they competed and references. The second function then matches the athletes with their link based on their first names.
```{r collect href, echo=TRUE, message=FALSE}
# all href in table with tableNr
hrefs_in_table = function(tableNr){
  links = url %>%
    read_html() %>%
    html_nodes('table.wikitable') %>%
    .[[tableNr]] %>%
    html_nodes('a') %>%
    html_attr('href') %>%
    paste0('https://en.wikipedia.org', .)
  (unique(links)) # remove duplicate entries
}

# add links column and match links to athlete
add_athlete_href = function(data_table, links){
  data_table$Link = NA
  for (idx in (1:length(data_table$Athlete))){
    athlete = data_table$Athlete[idx]
    name = strsplit(toString(athlete), " ")[[1]][1]
    matching_link = links[grep(name, links)]
    if(identical(matching_link, character(0))){next}
    data_table$Link[idx] = links[grep(name, links)]
  }
  return(data_table)
}
```
Sha'Carri Richardson's wiki link does not match the pattern, I add it explicitly. The tables format now looks like this:
```{r call href collecting functions, echo=FALSE, message=FALSE, warning=FALSE}
ATB_men_links = hrefs_in_table(2)
ATB_women_links = hrefs_in_table(3)
all_time_best_men = add_athlete_href(all_time_best_men, ATB_men_links)
all_time_best_women = add_athlete_href(all_time_best_women, ATB_women_links)

# one link didn't match pattern - insert missing
missing_link = which(all_time_best_women$Athlete == "Sha'Carri Richardson")
all_time_best_women$Link[missing_link] = "https://en.wikipedia.org/wiki/Sha%27Carri_Richardson"
glimpse(all_time_best_men)
```
Now lets collect the birthdays from their personal page. I take advantage of the typical info-box that's located on the top right of most wiki pages. However six of the entries aren't caught in the pattern (for various reasons) - I insert them in a post processing step.
```{r get birthdays, echo=TRUE, message=FALSE}
# web scrawl for birthdays to be able to compute age at record
add_athlete_birthday = function(data_table){
  data_table$Born = NA
  for (idx in (1:length(data_table$Athlete))){
    pg = read_html(data_table$Link[idx])
    info_box <- html_nodes(pg, xpath="//table[@class = 'infobox vcard']")
    bday_node <- html_nodes(info_box, xpath = "//span[@class = 'bday']")
    bday = html_text(bday_node)[1]
    data_table$Born[idx] = bday
  }
  return(data_table)
}
```
```{r call get birthdays and postprocessing, echo=FALSE, message=FALSE}
all_time_best_men = add_athlete_birthday(all_time_best_men)
all_time_best_women = add_athlete_birthday(all_time_best_women)

# hard-coding missing entries for female birthdates, where birthdate not accessible with the pattern
all_time_best_women$Born[c(4, 9, 11, 15, 21, 25)] = 
  c("1986-12-27", "1984-04-16",  "1957-04-15", "1966-03-27", "1972-06-12", "1966-11-19")
```
Next I convert the dates when the records where run to standard date format "YYYY-MM-DD". Then I can calculate the age of the athletes at the time they ran their best race.
```{r convert date format functions, echo=FALSE, message=FALSE, results="hide"}
# else some date conversions returned NA
lct <- Sys.getlocale("LC_TIME"); Sys.setlocale("LC_TIME", "C")

# convert date to yyyy-mm-dd
standardize_date = function(data_table){
  data_table[] = lapply(data_table, as.character)
  for (idx in (1:length(data_table$Date))){
    date = as.Date(data_table$Date[idx], format("%d %B %Y"))
    data_table$Date[idx] = toString(date)
  }
  return(data_table)
}

# add age at record: convert date of record into yyyy-mm-dd format and calc age.
calc_age_at_record = function(data_table){
  data_table[] = lapply(data_table, as.character)
  data_table$Age_at_record = NA
  for (idx in (1:length(data_table$Date))){
    date = as.Date(data_table$Date[idx], format("%d %B %Y"))
    data_table$Date[idx] = toString(date)
    data_table$Age_at_record[idx] = 
      as.numeric(difftime(date, as.Date(data_table$Born[idx]), units = "weeks"))/52.25
  }
  return(data_table)
}

all_time_best_men = calc_age_at_record(all_time_best_men)
all_time_best_women = calc_age_at_record(all_time_best_women)
```
Bringing the two tables together we can analyse the data. The overall mean age at peak performance of the top 25 women and men amounts to 25.8 years. The athletes gender appears to have no significant influence on the age the athletes set their records.
```{r GOAT table, echo=FALSE, message=FALSE}
# create common table (could do so earlier) and call functions once
all_time_best_men$Gender = "male"
all_time_best_women$Gender = "female"

# combine into one table, remove links column
GOAT = subset(rbind(all_time_best_men, all_time_best_women), select = -Link)

# analyze age at time of records
avg_age_f = mean(filter(GOAT, Gender == "female")$Age_at_record)
avg_age_m = mean(filter(GOAT, Gender == "male")$Age_at_record)
avg = mean(GOAT$Age_at_record)
stdev_f = sd(filter(GOAT, Gender == "female")$Age_at_record)
stdev_m = sd(filter(GOAT, Gender == "male")$Age_at_record)
stdev = sd(filter(GOAT, Gender == "female")$Age_at_record)
```
```{r box plotts, echo=FALSE, message=FALSE, warning=FALSE}
# boxplots of math by prog, with jittered data points
# install.packages("ggplot2")
library(ggplot2)

ggplot(data=GOAT, aes(x=Gender, y=Age_at_record)) +
  geom_boxplot() +
  geom_jitter(width=.035) 
# + theme(title = element_text("Age at record of all-time top 25 athletes by gender"), axis.title.y = element_text("Age at record"))
```  

The box plots show the data split in four quantiles (25% of the data per quantile): The middle line in the box is represented by the median, the box contains the upper and lower quantile, or the 50% of the data closest to the median. The whiskers visualize the remaining two, most extreme quantile groups. The age distribution looks quite similar between the genders. The respectable means are at 26.0 for female runners and 25.7 for males. The standard deviation for the female runners is 3.7 and for the males 3.4. My take is that I can not read a significant difference in prime age for athletic performances per genders from that data.

## Best month to be born a sprinter
This brings me to the final question I want to address: Is there a best time of the year to be born an 100 meters athlete? This question is based on a hypothesis I heard before. Namely, that kids born at the end of the year would be advantaged in sports, as they are always the oldest in their age groups. This of course assumes that the categories are drawn by the birth year and not current age. If the categories are given by the current age, as suggested by the '100 meters per age category' section, then maybe one could still find a trend if there are fixed seasons of competitions. For example such a season could be during months with mild weather for outdoors competition.

To analyze this I gather the birth months for junior (per age category record holders) and senior (all time top 25) in a table. The month values of senior athletes can be directly read from the birthdays. From the juniors I calculate the birth month based on their record date and age at record.

```{r best month, echo=FALSE, message=FALSE}
# install.packages("lubridate")
library(lubridate)

juniors_birth_month = select(rbind(per_age_boys, per_age_girls), Athlete, Date, Age)
juniors_birth_month = standardize_date(juniors_birth_month)

# set junior flag,  calc birthmonth, remove duplicates
juniors_birth_month$Month = NA
for (idx in (1:length(juniors_birth_month$Date))){
  date = as.Date(juniors_birth_month$Date[idx], format("%d %B %Y"))
  age_days = as.numeric(str_extract(strsplit(toString(juniors_birth_month$Age[idx]), ",")[[1]][2], "\\d+"))
  juniors_birth_month$Month[idx] = month(as.Date(juniors_birth_month$Date[idx])+(365-age_days))
}

juniors_just_month = unique(select(juniors_birth_month, Athlete, Month))
juniors_just_month$Category = "junior"

# calc age of young athletes and check if there's a trend in birthmonth
GOAT$Month = NA
for (idx in (1:length(GOAT$Born))){
  GOAT$Month[idx] = month(as.Date(GOAT$Born[idx]))
}
seniors_just_month = select(GOAT, Athlete, Month)
seniors_just_month$Category = "senior"

# common table
just_month = rbind(juniors_just_month, seniors_just_month)
glimpse(just_month)
```
Lets illustrate the data in a bar plot.
```{r month plot, echo=FALSE, message=FALSE}
# barplot birthmonth
tab = table(just_month$Category, just_month$Month)
barplot(tab, 
        legend.text = TRUE, xlab = "Birth month", ylab = "Athletes", args.legend = list(x = "topleft"))

# probability of getting this data with random birthdays
month_avg = mean(just_month$Month) # expected 6.5
month_var = var(just_month$Month) # expected 10.083
```  

For me it's hard to read any trend out of it by naked eye. It could follow a uniform distribution. This would indicate that the birth month has no influence on the chance to become a record setting athlete. To compare it with an uniform distribution lets compare the mean and variance to the expected mean and variance. We have a = 1, b = 12. The theoretical mean corresponds to (a+b)/2 = 6.5, the variance = ((b-a)^2)/b = 10.08333. The values we measure from the data seems to confirm a uniform distribution: mean is measured at 7.1 and variance 10.4. Given our data set of just 73 observations I believe this is comparable. I ran a maximum likelihood estimation algorithm to compare the generated estimators to a and b.
```{r MLE, echo=TRUE, message=FALSE}
# install.packages("ExtDist")
library(ExtDist)
len = length(just_month$Month)
mle_uniform = eUniform(X = just_month$Month, method = "unbiased.MLE")
print(mle_uniform)
```
From what I understand this estimated the most likely parameters a and b given the data and assuming an uniform distribution. With that I conclude that the birth months in record setting 100 meters athletes appear to follow the uniform distribution. Hence if there's a best month to be born a runner, it did not manifest in this data.